Portable Multi- and Many-Core Performance for Finite Difference Codes; Application to the Free-Surface Component of NEMO.

### 2017-05-26 nemolite2d_perf.tex
nemolite2d-paper: change to copernicus style

%% nemolite2d_perf.tex
%%
% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
% should be used if it is desired that the figures are to be displayed in
% draft mode.
%
\documentclass[journal]{IEEEtran}
%
% If IEEEtran.cls has not been installed into the LaTeX system files,
% manually specify the path to it like:
% \documentclass[journal]{../sty/IEEEtran}

% Also, note that IEEEtran.cls V1.7 and later provides a builtin
% \ifCLASSINFOpdf conditional that works the same way.
% When switching from latex to pdflatex and vice-versa, the compiler may
% have to be run twice to clear warning/error messages.

\usepackage{cite}

% *** GRAPHICS RELATED PACKAGES ***
%
\ifCLASSINFOpdf
  \usepackage[pdftex]{graphicx}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
\else
  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
  % will default to the driver specified in the system graphics.cfg if no
  % driver is specified.
  \usepackage[dvips]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../eps/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  \DeclareGraphicsExtensions{.eps}
\fi

% *** MATH PACKAGES ***
%
\usepackage[cmex10]{amsmath}
% A popular package from the American Mathematical Society that provides
% many useful and powerful commands for dealing with mathematics. If using
% it, be sure to load this package with the cmex10 option to ensure that
% only type 1 fonts will utilized at all point sizes. Without this option,
% it is possible that some math symbols, particularly those within
% footnotes, will be rendered in bitmap form which will result in a
% document that can not be IEEE Xplore compliant!
%
% Also, note that the amsmath package sets \interdisplaylinepenalty to 10000
% thus preventing page breaks from occurring within multiline equations. Use:
%\interdisplaylinepenalty=2500
% after loading amsmath to restore such page breaks as IEEEtran.cls normally
% does. amsmath.sty is already installed on most LaTeX systems. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/required/amslatex/math/

% *** SPECIALIZED LIST PACKAGES ***
%
%\usepackage{algorithmic}
% algorithmic.sty was written by Peter Williams and Rogerio Brito.
% This package provides an algorithmic environment fo describing algorithms.
% You can use the algorithmic environment in-text or within a figure
% environment to provide for a floating algorithm. Do NOT use the algorithm
% floating environment provided by algorithm.sty (by the same authors) or
% algorithm2e.sty (by Christophe Fiorio) as IEEE does not use dedicated
% algorithm float types and packages that provide these will not provide
% correct IEEE style captions. The latest version and documentation of
% algorithmic.sty can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithms/
% There is also a support site at:
% http://algorithms.berlios.de/index.html
% Also of interest may be the (relatively newer and more customizable)
% algorithmicx.sty package by Szasz Janos:
% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithmicx/

% *** ALIGNMENT PACKAGES ***
%
%\usepackage{array}
% Frank Mittelbach's and David Carlisle's array.sty patches and improves
% the standard LaTeX2e array and tabular environments to provide better
% appearance and additional user controls. As the default LaTeX2e table
% generation code is lacking to the point of almost being broken with
% respect to the quality of the end results, all users are strongly
% advised to use an enhanced (at the very least that provided by array.sty)
% set of table tools. array.sty is already installed on most systems. The
% latest version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/required/tools/

% IEEEtran contains the IEEEeqnarray family of commands that can be used to
% generate multiline equations as well as matrices, tables, etc., of high
% quality.

% *** SUBFIGURE PACKAGES ***
%\ifCLASSOPTIONcompsoc
%  \usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
%\else
%  \usepackage[caption=false,font=footnotesize]{subfig}
%\fi
% subfig.sty, written by Steven Douglas Cochran, is the modern replacement
% for subfigure.sty, the latter of which is no longer maintained and is
% incompatible with some LaTeX packages including fixltx2e. However,
% subfig.sty requires and automatically loads Axel Sommerfeldt's caption.sty
% which will override IEEEtran.cls' handling of captions and this will result
% in non-IEEE style figure/table captions. To prevent this problem, be sure
% and invoke subfig.sty's "caption=false" package option (available since
% subfig.sty version 1.3, 2005/06/28) as this is will preserve IEEEtran.cls
% handling of captions.
% Note that the Computer Society format requires a larger sans serif font
% than the serif footnote size font used in traditional IEEE formatting
% and thus the need to invoke different subfig.sty package options depending
% on whether compsoc mode has been enabled.
%
% The latest version and documentation of subfig.sty can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/subfig/

% *** FLOAT PACKAGES ***
%
%\usepackage{fixltx2e}
% fixltx2e, the successor to the earlier fix2col.sty, was written by
% Frank Mittelbach and David Carlisle. This package corrects a few problems
% in the LaTeX2e kernel, the most notable of which is that in current
% LaTeX2e releases, the ordering of single and double column floats is not
% guaranteed to be preserved. Thus, an unpatched LaTeX2e can allow a
% single column figure to be placed prior to an earlier double column
% figure. The latest version and documentation can be found at:
% http://www.ctan.org/tex-archive/macros/latex/base/

%\usepackage{stfloats}
% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
% the ability to do double column floats at the bottom of the page as well
% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
% LaTeX2e). It also provides a command:
%\fnbelowfloat
% to enable the placement of footnotes below bottom floats (the standard
% LaTeX2e kernel puts them above bottom floats). This is an invasive package
% which rewrites many portions of the LaTeX2e float routines. It may not work
% with other packages that modify the LaTeX2e float routines. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/sttools/
% Do not use the stfloats baselinefloat ability as IEEE does not allow
% \baselineskip to stretch. Authors submitting work to the IEEE should note
% that IEEE rarely uses double column equations and that authors should try
% to avoid such use. Do not be tempted to use the cuted.sty or midfloat.sty
% packages (also by Sigitas Tolusis) as IEEE does not format its papers in
% such ways.
% Do not attempt to use stfloats with fixltx2e as they are incompatible.
% Instead, use Morten Hogholm'a dblfloatfix which combines the features
% of both fixltx2e and stfloats:
%
\usepackage{dblfloatfix}

%\ifCLASSOPTIONcaptionsoff
%  \usepackage[nomarkers]{endfloat}
% \let\MYoriglatexcaption\caption
% \renewcommand{\caption}[2][\relax]{\MYoriglatexcaption[#2]{#2}}
%\fi
% endfloat.sty was written by James Darrell McCauley, Jeff Goldberg and 
% Axel Sommerfeldt. This package may be useful when used in conjunction with 
% IEEEtran.cls'  captionsoff option. Some IEEE journals/societies require that
% submissions have lists of figures/tables at the end of the paper and that
% figures/tables without any captions are placed on a page by themselves at
% the end of the document. If needed, the draftcls IEEEtran class option or
% \CLASSINPUTbaselinestretch interface can be used to increase the line
% spacing as well. Be sure and use the nomarkers option of endfloat to
% prevent endfloat from "marking" where the figures would have been placed
% in the text. The two hack lines of code above are a slight modification of
% that suggested by in the endfloat docs (section 8.4.1) to ensure that
% the full captions always appear in the list of figures/tables - even if
% the user used the short optional argument of \caption[]{}.
% IEEE papers do not typically make use of \caption[]'s optional argument,
% so this should not be an issue. A similar trick can be used to disable
% captions of packages such as subfig.sty that lack options to turn off
% the subcaptions:
% For subfig.sty:
% \let\MYorigsubfloat\subfloat
% \renewcommand{\subfloat}[2][\relax]{\MYorigsubfloat[]{#2}}
% However, the above trick will not work if both optional arguments of
% the \subfloat command are used. Furthermore, there needs to be a
% description of each subfigure *somewhere* and endfloat does not add
% subfigure captions to its list of figures. Thus, the best approach is to
% avoid the use of subfigure captions (many IEEE journals avoid them anyway)
% and instead reference/explain all the subfigures within the main caption.
% The latest version of endfloat.sty and its documentation can obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/endfloat/
%
% The IEEEtran \ifCLASSOPTIONcaptionsoff conditional can also be used
% later in the document, say, to conditionally put the References on a 
% page by themselves.

\usepackage{url}

% correct bad hyphenation here
\hyphenation{op-tical net-works semi-conduc-tor}

\newlength{\picwidth}
\setlength{\picwidth}{85mm}

\newcommand{\psykal}{{PS}y{KA}l}

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{Portable Multi- and Many-Core Performance for Finite Difference Codes;
  Application to the Free-Surface Component of NEMO.}
%
% author names and IEEE memberships
% use \thanks{} to gain access to the first footnote area
% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
% was not built to handle multiple paragraphs
%

\author{A.~R.~Porter, J.~Appleyard, M.~Ashworth, R.~W.~Ford, J.~Holt, H.~Liu and G.~D.~Riley% <-this % stops a space
\thanks{M.~Ashworth, R.~W.~Ford and A.~R.~Porter are with the Science and Technology Facilities Council, UK.}% <-this % stops a space
\thanks{J.~Appleyard is with NVIDIA.}%
\thanks{H.~Liu and J.~Holt are with the National Oceanography Centre, Liverpool, UK.}%
\thanks{G.~Riley is with the University of Manchester, Manchester, UK.}% <-this % stops a space
}

% note the % following the last \IEEEmembership and also \thanks - 
% these prevent an unwanted space from occurring between the last author name
% and the end of the author line. i.e., if you had this:
% 
% \author{....lastname \thanks{...} \thanks{...} }
%                     ^------------^------------^----Do not want these spaces!
%
% a space would be appended to the last name and could cause every name on that
% line to be shifted left slightly. This is one of those "LaTeX things". For
% instance, "\textbf{A} \textbf{B}" will typeset as "A B" not "AB". To get
% "AB" then you have to do: "\textbf{A}\textbf{B}"
% \thanks is no different in this regard, so shield the last } of each \thanks
% that ends a line with a % and do not let a space in before the next \thanks.
% Spaces after \IEEEmembership other than the last one are OK (and needed) as
% you are supposed to have spaces between the names.

% The paper headers
\markboth{Journal of \LaTeX\ Class Files,~Vol.~11, No.~4, December~2012}%
{Shell \MakeLowercase{\textit{et al.}}: Bare Demo of IEEEtran.cls for Journals}
% The only time the second header will appear is for the odd numbered pages
% after the title page when using the twoside option.
% 
% *** Note that you probably will NOT want to include the author's ***
% *** name in the headers of peer review papers.                   ***
% You can use \ifCLASSOPTIONpeerreview for conditional compilation here if
% you desire.

% If you want to put a publisher's ID mark on the page you can do it like
% this:
%\IEEEpubid{0000--0000/00\$00.00~\copyright~2012 IEEE}
% Remember, if you use this you must call \IEEEpubidadjcol in the second
% column for its text to clear the IEEEpubid mark.

% make the title area
\maketitle

@@

We present an approach which we call \psykal\ that is designed to achieve
portable performance for parallel, finite-difference Ocean models.  In
\psykal\ the code related to the underlying science is formally
separated from code related to parallelisation and single-core
optimisations. This separation of concerns allows scientists to code
their science independently of the underlying hardware architecture
and for optimisation specialists to be able to tailor the code for a
particular machine independently of the science code. We have taken
the free-surface part of the NEMO ocean model and created a new,
shallow-water model named NEMOLite2D. In doing this we have a code
which is of a manageable size and yet which incorporates elements of
full ocean models (input/output, boundary conditions, \textit{etc.}).
We have then manually constructed a \psykal\ version of this code and
investigated the transformations that must be applied to the
middle/PSy layer in order to achieve good performance, both serial and
parallel. We have produced versions of the PSy layer parallelised with
both OpenMP and OpenACC; in both cases we were able to leave the
natural-science parts of the code unchanged while achieving good
performance on both multi-core CPUs and GPUs. In quantifying whether
or not the obtained performance is `good' we also consider the
limitations of the basic roofline model and improve on it
by generating kernel-specific CPU ceilings.

@@

% Note that keywords are not normally used for peerreview papers.
\begin{IEEEkeywords}
Performance, Code-generation, Finite-difference, NEMO, OpenMP, OpenACC, roofline
\end{IEEEkeywords}

% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi
%
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
\IEEEpeerreviewmaketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps.
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
% 
% Here we have the typical use of a "T" for an initial drop letter
% and "HE" in caps to complete the first word.
\IEEEPARstart{T}{he} challenge presented to the developers of
scientific software by the drive towards Exascale computing is
considerable. With power consumption becoming the overriding design
constraint, CPU clock speeds are falling and the complex,
multi-purpose compute core is being replaced by multiple, simpler
cores. This philosophy can be seen at work in the rise of so-called
accelerator-based machines in the Top 500 List~\cite{top500} of
supercomputers: four of the top-ten machines in the June 2016 list
make use of many-core processors (Intel Xeon Phi, NVIDIA GPU or NRCPC
SW26010). Two of the remaining six machines are IBM BlueGene/Qs, the
CPU of which has hardware support for running 64 threads.

@@

In a previous paper~\cite{shallow_psykal} we introduced a possible
approach to tackling this problem which we term \psykal\ (discussed
below). In that work we considered the implications for serial
performance of the extensive code re-structuring required by the
approach when applied to the `Shallow' shallow-water model
(https://puma.nerc.ac.uk/trac/GOcean). We found that although the
re-structuring did initially incur a sizeable performance penalty, it
was possible to transform the resulting code to recover performance
(for a variety of CPU/compiler combinations) while obeying the
\psykal\ separation of concerns. In this work we move to looking at
portable parallel performance within the \psykal\ approach.

% needed in second column of first page if using \IEEEpubid
%\IEEEpubidadjcol

\subsection{The \psykal\ Approach}

The \psykal\ approach attempts to address the problems described in
the previous section. It separates code into three layers; the
Algorithm layer, the PSy layer and the Kernel layer. The approach has
been developed in the GungHo project~\cite{GungHo}, which is creating
a new Dynamical core for the UK Met Office, and its design has been
influenced by earlier work on OP2~\cite{OP2, PYOP2}.

In common with OP2, the \psykal\ approach separates out the science
code and the performance-related code into distinct layers. The calls
that specify parallelism in both approaches are similar in terms of
where they are placed in the code and in their semantics. However, the
\psykal\ approach supports the specification of more than one kernel
in a parallel region of code, compared with one for OP2, giving more
scope for optimisation. In addition, the metadata describing a kernel
is included with the kernel code in the \psykal\ approach whereas it
is provided as a part of the kernel call in OP2.

While the \psykal\ approach is general, we are currently applying it to
Atmosphere and Ocean models written in Fortran where domain
decomposition is typically performed in the latitude-longitude
dimension, leaving columns of elements on each domain-decomposed
partition.

@@

Since any real oceanographic computational model must output results,
we ensure that any \psykal\ version of NEMOLite2D retains the Input/Output
capability of the original. This aids in limiting the optimisations
that can be performed on the \psykal\ version to those that should also
be applicable to full oceanographic models. Note that although we
retain the I/O functionality, all of the results presented in this work
carefully exclude the effects of I/O since it is compute performance
that interests us here.

In order to maximise the flexibility (and thus potential for
optimisation) of the \psykal\ version of NEMOLite2D, we made the
kernels as fine-grained as possible. This resulted in eight distinct
kernels, each of which updates a single field at a single point (since
we have chosen to use point-wise kernels). With a little bit of
tidying/re-structuring, we found it was possible to express the
contents of the main time-stepping loop as a single invoke (a call to
the PSy layer) and a call to the I/O system
(Figure~\ref{FIG_psykal_nemolite2d_structure}). The single invoke
gives us a single PSy-layer routine which consists of applying each of
the kernels to all of the points requiring an update on the model
mesh. In its basic, unoptimised (`vanilla') form, this PSy-layer
routine then contains a doubly-nested loop (over the two dimensions of
the model grid) around each kernel call.

\begin{figure}
\centering
\begin{verbatim}
DO istep = 1, nsteps
  call invoke( continuity(ssha_t, ...), &
               momentum_u(ua, un, ...), &
               ...,                     &
               next_sshu(sshn_u, ...),  &
               next_sshv(sshn_v, ...) )
  call model_write(istep, sshn_t, un, vn)
END DO
\end{verbatim}
\caption{A schematic of the top-level of the \psykal\ version of the
 NEMOLite2D code.}
\label{FIG_psykal_nemolite2d_structure}
\end{figure}

@@

Before applying any code transformations, we first benchmark the
original, serial version of the code. We also benchmark the vanilla,
unoptimised version after it has been re-structured following the
\psykal\ approach. In addition to this benchmarking, we profile these
versions of the code at the algorithm level (using a high-level timing
API). The resulting profiles are given in
Table~\ref{TABLE_profile}. The Momentum section dominates the profile
of both versions of the code, accounting for around 70--80\% of the
wall-clock time spent doing time-stepping. It is also the key section
when we consider the performance loss when moving from the original to
the \psykal\ version of the code; it slows down by a factor of
two. Although less significant in terms of absolute time, the
Continuity section is also dramatically slower in the \psykal\ version,
this time by a factor of three. In contrast, the performance of the
Time-update and Boundary Condition regions are not significantly
affected by the move to the \psykal\ version.

\begin{table}[!t]
\caption{The performance profile of the original and \psykal\ versions
  of NEMOLite2D on the Intel Ivy Bridge CPU (for 2000 time-steps of the
  $128^{2}$ domain and the Intel compiler).}
\label{TABLE_profile}
\centering
\begin{tabular}{l|c|c|c|c|c|c}
\hline
Section & \multicolumn{2}{c|}{Original} & \multicolumn{2}{c|}{Vanilla \psykal} &
\multicolumn{2}{c}{Final \psykal} \\
        & Time (s) & \%  & Time (s) & \%  & Time(s) & \% \\
\hline
Momentum    & 1.98  & 72.6  & 4.05  & 79.5 & 2.09 & 75.3 \\
Time-update & 0.40  & 14.6  & 0.41  & 8.1  & 0.29 & 10.6 \\
BCs         & 0.25  & 9.1   & 0.29  & 5.7  & 0.27 & 9.9  \\
Continuity  & 0.10  & 3.7   & 0.33  & 6.6  & 0.11 & 4.1  \\
\hline
\end{tabular}
\end{table}

Beginning with the vanilla \psykal\ version, we then apply a series of
code transformations while obeying the \psykal\ separation of concerns,
{\it i.e.} optimisation is restricted to the middle, {PS}y layer and
leaves the kernel and algorithm layers unchanged. The aim of these
optimisations is to recover, as much as is possible, the performance
of the original version of the code. The transformations we have
performed and the reasons for them are described in the following sections.

@@

The profiling data in Table~\ref{TABLE_profile} shows that it is the
Momentum section that accounts for the bulk of the model run-time.  We
therefore chose to attempt to optimise this section first. In-keeping
with the \psykal\ approach, we are only permitted to optimise the middle
(PSy) layer which, for this section comprises calls to two kernels,
one for each of the $x$ and $y$ components of momentum. These kernels
are relatively large; each comprises roughly 85 lines of Fortran
executable statements.

@@

It turns out that the Cray-compiled binaries of both the original and
\psykal\ versions of NEMOLite2D perform considerably less well than
their Intel-compiled counterparts. Comparison of the diagnostic output
from each of the compilers revealed that while the Intel compiler was
happy to vectorise the Momentum loops, the Cray compiler was choosing
not to:
\begin{verbatim}
   99.          do ji = 2, M-1, 1
 A loop starting at line 99 was blocked 
 with block size 256.

@@

Having optimised the Momentum section as much as permitted by the
\psykal\ approach, we turn our attention to the three remaining
sections of the code. The profile data in Table~\ref{TABLE_profile}
shows that these regions are all comparable in terms of cost. What is
striking however, is that the cost of the Continuity section increases
by more than a factor of three in moving to the \psykal\ version of the
code.

@@

For this part of the work we began with the optimal \psykal\ version of
the code, as obtained after applying the various transformations
described in the previous section. As with the transformations of the
serial code, our purpose here is to determine the functionality
required of a tool that seeks to generate the PSy layer.

@@

Moving now the to the \psykal\ version of NEMOLite2D,
Figure~\ref{FIG_best_psykal_perf_summary} plots the performance of the
fastest \psykal\ version for each of the compiler/CPU combinations.
While the Intel compiler still produces the best-performing binary,
the Cray-compiled binary is now a very close second. In fact, the
performance of both the Gnu- and Cray-compiled \psykal\ versions is
generally significantly greater than that of their respective original
versions. We also note that best absolute performance (in terms of
grid points processed per second) with any compiler is obtained with
the $256^2$ domain.

\begin{figure}[!t]
\centering
\includegraphics[width=7.5cm]{best_psykal_summary}
\caption{Summary of the best performance achieved by any \psykal\ 
version of NEMOLite2D with each of the compilers and CPUs under 
consideration.}
\label{FIG_best_psykal_perf_summary}
\end{figure}

Figure~\ref{FIG_slowdown_summary} plots the percentage difference
between the performance of the original and the best \psykal\ versions
of NEMOLite2D for each compiler/CPU combination. This shows that it is
only the Intel-compiled binary running the $64^2$ domain that is
slower with the \psykal\ version of the code (and then only by some
3\%). For all other points in the space, the optimised \psykal\ version
of the code performs better. The results for the Cray compiler are
however somewhat sqewed by the fact that it did not SIMD vectorise key
parts of the original version (see below).

\begin{figure}[!t]
\centering
\includegraphics[width=2.8in]{slowdown_summary}
\caption{Comparison of the performance of the best \psykal\
version with that of the original version of the code. A negative value 
indicates that the \psykal\ version is slower than the original.}
\label{FIG_slowdown_summary}
\end{figure}

>>>>

%% 2-column papers and discussion papers
\documentclass[gmd, manuscript]{copernicus}

%% \usepackage commands included in the copernicus.cls:
%\usepackage[german, english]{babel}
%\usepackage{tabularx}
%\usepackage{cancel}
%\usepackage{multirow}
%\usepackage{supertabular}
%\usepackage{algorithmic}
%\usepackage{algorithm}
%\usepackage{amsthm}
%\usepackage{float}
%\usepackage{subfig}
%\usepackage{rotating}

%\usepackage{cite}
%\usepackage{url}

\newlength{\picwidth}
\setlength{\picwidth}{83mm}

\begin{document}

\title{Portable Multi- and Many-Core Performance for Finite Difference Codes;
  Application to the Free-Surface Component of NEMO.}

\Author[1]{Andrew}{Porter}
\Author[2]{Jeremy}{Appleyard}
\Author[1]{Mike}{Ashworth}
\Author[1]{Rupert}{Ford}
\Author[3]{Jason}{Holt}
\Author[3]{Hedong}{Liu}
\Author[4]{Graham}{Riley}

\affil[1]{Science and Technology Facilities Council, Daresbury Laboratory, UK}
\affil[2]{NVIDIA}
\affil[3]{National Oceanography Centre, Liverpool, UK}
\affil[4]{University of Manchester, Manchester, UK}

\runningtitle{Portable Multi- and Many-Core Performance for Finite Difference Codes}

\runningauthor{A.~R.~Porter et al}

\correspondence{Andrew Porter (andrew.porter@stfc.ac.uk)}

\received{}
\pubdiscuss{} %% only important for two-stage journals
\revised{}
\accepted{}
\published{}

\firstpage{1}
\maketitle

@@

We present an approach which we call {PS}y{KA}l that is designed to achieve
portable performance for parallel, finite-difference Ocean models.  In
{PS}y{KA}l the code related to the underlying science is formally
separated from code related to parallelisation and single-core
optimisations. This separation of concerns allows scientists to code
their science independently of the underlying hardware architecture
and for optimisation specialists to be able to tailor the code for a
particular machine independently of the science code. We have taken
the free-surface part of the NEMO ocean model and created a new,
shallow-water model named NEMOLite2D. In doing this we have a code
which is of a manageable size and yet which incorporates elements of
full ocean models (input/output, boundary conditions, \textit{etc.}).
We have then manually constructed a {PS}y{KA}l version of this code and
investigated the transformations that must be applied to the
middle/PSy layer in order to achieve good performance, both serial and
parallel. We have produced versions of the PSy layer parallelised with
both OpenMP and OpenACC; in both cases we were able to leave the
natural-science parts of the code unchanged while achieving good
performance on both multi-core CPUs and GPUs. In quantifying whether
or not the obtained performance is `good' we also consider the
limitations of the basic roofline model and improve on it
by generating kernel-specific CPU ceilings.

@@

\copyrightstatement{Copyright Science and Technology Facilities Council, 2017}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\introduction

The challenge presented to the developers of scientific software by
the drive towards Exascale computing is considerable. With power
consumption becoming the overriding design constraint, CPU clock
speeds are falling and the complex, multi-purpose compute core is
being replaced by multiple, simpler cores. This philosophy can be seen
at work in the rise of so-called accelerator-based machines in the Top
500 List~\cite{top500} of supercomputers: four of the top-ten machines
in the June 2016 list make use of many-core processors (Intel Xeon
Phi, NVIDIA GPU or NRCPC SW26010). Two of the remaining six machines
are IBM BlueGene/Qs, the CPU of which has hardware support for running
64 threads.

@@

In a previous paper~\cite{shallow_psykal} we introduced a possible
approach to tackling this problem which we term {PS}y{KA}l (discussed
below). In that work we considered the implications for serial
performance of the extensive code re-structuring required by the
approach when applied to the `Shallow' shallow-water model
(https://puma.nerc.ac.uk/trac/GOcean). We found that although the
re-structuring did initially incur a sizeable performance penalty, it
was possible to transform the resulting code to recover performance
(for a variety of CPU/compiler combinations) while obeying the
{PS}y{KA}l separation of concerns. In this work we move to looking at
portable parallel performance within the {PS}y{KA}l approach.

\subsection{The {PS}y{KA}l Approach}

The {PS}y{KA}l approach attempts to address the problems described in
the previous section. It separates code into three layers; the
Algorithm layer, the PSy layer and the Kernel layer. The approach has
been developed in the GungHo project~\cite{GungHo}, which is creating
a new Dynamical core for the UK Met Office, and its design has been
influenced by earlier work on OP2~\cite{OP2, PYOP2}.

In common with OP2, the {PS}y{KA}l approach separates out the science
code and the performance-related code into distinct layers. The calls
that specify parallelism in both approaches are similar in terms of
where they are placed in the code and in their semantics. However, the
{PS}y{KA}l approach supports the specification of more than one kernel
in a parallel region of code, compared with one for OP2, giving more
scope for optimisation. In addition, the metadata describing a kernel
is included with the kernel code in the {PS}y{KA}l approach whereas it
is provided as a part of the kernel call in OP2.

While the {PS}y{KA}l approach is general, we are currently applying it to
Atmosphere and Ocean models written in Fortran where domain
decomposition is typically performed in the latitude-longitude
dimension, leaving columns of elements on each domain-decomposed
partition.

@@

Since any real oceanographic computational model must output results,
we ensure that any {PS}y{KA}l version of NEMOLite2D retains the Input/Output
capability of the original. This aids in limiting the optimisations
that can be performed on the {PS}y{KA}l version to those that should also
be applicable to full oceanographic models. Note that although we
retain the I/O functionality, all of the results presented in this work
carefully exclude the effects of I/O since it is compute performance
that interests us here.

In order to maximise the flexibility (and thus potential for
optimisation) of the {PS}y{KA}l version of NEMOLite2D, we made the
kernels as fine-grained as possible. This resulted in eight distinct
kernels, each of which updates a single field at a single point (since
we have chosen to use point-wise kernels). With a little bit of
tidying/re-structuring, we found it was possible to express the
contents of the main time-stepping loop as a single invoke (a call to
the PSy layer) and a call to the I/O system
(Figure~\ref{FIG_psykal_nemolite2d_structure}). The single invoke
gives us a single PSy-layer routine which consists of applying each of
the kernels to all of the points requiring an update on the model
mesh. In its basic, unoptimised (`vanilla') form, this PSy-layer
routine then contains a doubly-nested loop (over the two dimensions of
the model grid) around each kernel call.

\begin{figure}
\centering
\begin{verbatim}
DO istep = 1, nsteps
  call invoke( continuity(ssha_t, ...), &
               momentum_u(ua, un, ...), &
               ...,                     &
               next_sshu(sshn_u, ...),  &
               next_sshv(sshn_v, ...) )
  call model_write(istep, sshn_t, un, vn)
END DO
\end{verbatim}
\caption{A schematic of the top-level of the {PS}y{KA}l version of the
 NEMOLite2D code.}
\label{FIG_psykal_nemolite2d_structure}
\end{figure}

@@

Before applying any code transformations, we first benchmark the
original, serial version of the code. We also benchmark the vanilla,
unoptimised version after it has been re-structured following the
{PS}y{KA}l approach. In addition to this benchmarking, we profile these
versions of the code at the algorithm level (using a high-level timing
API). The resulting profiles are given in
Table~\ref{TABLE_profile}. The Momentum section dominates the profile
of both versions of the code, accounting for around 70--80\% of the
wall-clock time spent doing time-stepping. It is also the key section
when we consider the performance loss when moving from the original to
the {PS}y{KA}l version of the code; it slows down by a factor of
two. Although less significant in terms of absolute time, the
Continuity section is also dramatically slower in the {PS}y{KA}l version,
this time by a factor of three. In contrast, the performance of the
Time-update and Boundary Condition regions are not significantly
affected by the move to the {PS}y{KA}l version.

\begin{table}[!t]
\caption{The performance profile of the original and {PS}y{KA}l versions
  of NEMOLite2D on the Intel Ivy Bridge CPU (for 2000 time-steps of the
  $128^{2}$ domain and the Intel compiler).}
\label{TABLE_profile}
\centering
\begin{tabular}{l|c|c|c|c|c|c}
\hline
Section & \multicolumn{2}{c|}{Original} & \multicolumn{2}{c|}{Vanilla {PS}y{KA}l} &
\multicolumn{2}{c}{Final {PS}y{KA}l} \\
        & Time (s) & \%  & Time (s) & \%  & Time(s) & \% \\
\hline
Momentum    & 1.98  & 72.6  & 4.05  & 79.5 & 2.09 & 75.3 \\
Time-update & 0.40  & 14.6  & 0.41  & 8.1  & 0.29 & 10.6 \\
BCs         & 0.25  & 9.1   & 0.29  & 5.7  & 0.27 & 9.9  \\
Continuity  & 0.10  & 3.7   & 0.33  & 6.6  & 0.11 & 4.1  \\
\hline
\end{tabular}
\end{table}

Beginning with the vanilla {PS}y{KA}l version, we then apply a series of
code transformations while obeying the {PS}y{KA}l separation of concerns,
{\it i.e.} optimisation is restricted to the middle, {PS}y layer and
leaves the kernel and algorithm layers unchanged. The aim of these
optimisations is to recover, as much as is possible, the performance
of the original version of the code. The transformations we have
performed and the reasons for them are described in the following sections.

@@

The profiling data in Table~\ref{TABLE_profile} shows that it is the
Momentum section that accounts for the bulk of the model run-time.  We
therefore chose to attempt to optimise this section first. In-keeping
with the {PS}y{KA}l approach, we are only permitted to optimise the middle
(PSy) layer which, for this section comprises calls to two kernels,
one for each of the $x$ and $y$ components of momentum. These kernels
are relatively large; each comprises roughly 85 lines of Fortran
executable statements.

@@

It turns out that the Cray-compiled binaries of both the original and
{PS}y{KA}l versions of NEMOLite2D perform considerably less well than
their Intel-compiled counterparts. Comparison of the diagnostic output
from each of the compilers revealed that while the Intel compiler was
happy to vectorise the Momentum loops, the Cray compiler was choosing
not to:
\begin{verbatim}
  99. do ji = 2, M-1, 1
 A loop starting at line 99 was blocked 
 with block size 256.

@@

Having optimised the Momentum section as much as permitted by the
{PS}y{KA}l approach, we turn our attention to the three remaining
sections of the code. The profile data in Table~\ref{TABLE_profile}
shows that these regions are all comparable in terms of cost. What is
striking however, is that the cost of the Continuity section increases
by more than a factor of three in moving to the {PS}y{KA}l version of the
code.

@@

For this part of the work we began with the optimal {PS}y{KA}l version of
the code, as obtained after applying the various transformations
described in the previous section. As with the transformations of the
serial code, our purpose here is to determine the functionality
required of a tool that seeks to generate the PSy layer.

@@

Moving now the to the {PS}y{KA}l version of NEMOLite2D,
Figure~\ref{FIG_best_psykal_perf_summary} plots the performance of the
fastest {PS}y{KA}l version for each of the compiler/CPU combinations.
While the Intel compiler still produces the best-performing binary,
the Cray-compiled binary is now a very close second. In fact, the
performance of both the Gnu- and Cray-compiled {PS}y{KA}l versions is
generally significantly greater than that of their respective original
versions. We also note that best absolute performance (in terms of
grid points processed per second) with any compiler is obtained with
the $256^2$ domain.

\begin{figure}[!t]
\centering
\includegraphics[width=7.5cm]{best_psykal_summary}
\caption{Summary of the best performance achieved by any {PS}y{KA}l 
version of NEMOLite2D with each of the compilers and CPUs under 
consideration.}
\label{FIG_best_psykal_perf_summary}
\end{figure}

Figure~\ref{FIG_slowdown_summary} plots the percentage difference
between the performance of the original and the best {PS}y{KA}l versions
of NEMOLite2D for each compiler/CPU combination. This shows that it is
only the Intel-compiled binary running the $64^2$ domain that is
slower with the {PS}y{KA}l version of the code (and then only by some
3\%). For all other points in the space, the optimised {PS}y{KA}l version
of the code performs better. The results for the Cray compiler are
however somewhat sqewed by the fact that it did not SIMD vectorise key
parts of the original version (see below).

\begin{figure}[!t]
\centering
\includegraphics[width=2.8in]{slowdown_summary}
\caption{Comparison of the performance of the best {PS}y{KA}l
version with that of the original version of the code. A negative value 
indicates that the {PS}y{KA}l version is slower than the original.}
\label{FIG_slowdown_summary}
\end{figure}
